---
title: "Analysis of Lead Exposure and IQ Levels"
author: "Data Analysis Report"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

# Load required packages
library(tidyverse)
library(knitr)
library(readr)
```

## Introduction

This report analyzes the relationship between proximity to a lead-emitting ore smelter and IQ scores in children from El Paso. The data comes from a CDC-led research team investigating the effects of low-level lead absorption on neurological function in children aged 3-15 years.

## Data Loading and Initial Processing

```{r load_data}
# Create the dataset
data <- read_csv("lead-iq-01.csv")
# Calculate initial summary statistics

means_by_location <- data %>%
  group_by(Smelter) %>%
  summarise(
    mean_iq = mean(IQ),
    sd_iq = sd(IQ),
    n = n()
  )
```

## Initial Analysis

### Visualization of IQ Levels by Location

```{r initial_plot, fig.cap="Distribution of IQ Scores by Proximity to Smelter (Before Correction)"}
ggplot(data, aes(x = Smelter, y = IQ)) +
  geom_boxplot(fill = "lightblue", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkblue") +
  theme_minimal() +
  labs(
    title = "Distribution of IQ Scores by Proximity to Smelter",
    x = "Distance from Smelter",
    y = "IQ Score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )
```

The initial boxplot reveals an unusual observation in the "Far" group with an IQ score of 999, which appears to be a data entry error.

### Initial Summary Statistics

```{r initial_summary}
  kable(means_by_location,
    col.names = c("Location", "Mean IQ", "SD", "N"),
    digits = 2,
    caption = "Summary Statistics of IQ Scores by Location (Before Correction)"
  )


```

## Corrected Analysis

After consulting with the primary investigator, we learned that the IQ value of 999 should be 99. Let's correct this error and reanalyze the data.

```{r update_data}
# Fix the incorrect value
data$IQ[data$IQ == 999] <- 99

# Recalculate means
updated_means <- data %>%
  group_by(Smelter) %>%
  summarise(
    mean_iq = mean(IQ),
    sd_iq = sd(IQ),
    n = n()
  )


```

### Updated Visualization

```{r corrected_plot, fig.cap="Distribution of IQ Scores by Proximity to Smelter (After Correction)"}
ggplot(data, aes(x = Smelter, y = IQ)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkgreen") +
  theme_minimal() +
  labs(
    title = "Distribution of IQ Scores by Proximity to Smelter (Corrected Data)",
    x = "Distance from Smelter",
    y = "IQ Score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )

```

### Updated Summary Statistics

```{r corrected_summary}
kable(updated_means,
      caption = "Summary Statistics of IQ Scores by Location (After Correction)",
      col.names = c("Location", "Mean IQ", "SD", "N"),
      digits = 2)
```

## Results and Discussion

After correcting the data entry error, we can see that children living far from the smelter have a mean IQ of `r round(updated_means$mean_iq[updated_means$Smelter == "Far"], 2)` (SD = `r round(updated_means$sd_iq[updated_means$Smelter == "Far"], 2)`), while those living near the smelter have a mean IQ of `r round(updated_means$mean_iq[updated_means$Smelter == "Near"], 2)` (SD = `r round(updated_means$sd_iq[updated_means$Smelter == "Near"], 2)`).

The corrected visualization and summary statistics show a clear difference in IQ scores between the two groups, with children living far from the smelter showing generally higher IQ scores than those living near the smelter. This suggests a possible relationship between proximity to the lead-emitting smelter and cognitive development as measured by IQ scores.